### 2014-10-15 Abstract.tex
all the changes

We provide a modular proof of a high-performance implementation of Lamport's
sequential consistency in a system with distributed memory. Unlike
Lamport's reference operational model, a high-performance implementation
processes many memory accesses concurrently. Processors employ
\emph{speculative execution} to generate many concurrent loads to pipeline the
memory system effectively. We define what it means for a speculative processor
to be correct independently of the memory system. We then define a memory
system with caches and show that our cache-coherence protocol preserves
\emph{store atomicity} (independently of the processor), a property that all
reasonable memories have in common.  We offer the first formal proof that when
an atomic memory is connected to $n$ speculative processors, the system as a
whole is sequentially consistent. The proof is composed modularly from proofs
of the individual components. Additionally, the cache-coherent memory we
present in this paper is the first fully verified system that supports
arbitrary cache hierarchies. We describe all our components using labeled
transition systems and machine-verify all our proofs using the Coq proof
assistant.

>>>>

We provide a framework for modular verification of multi-processor hardware
systems with respect to memory model specification. In this framework, we make
use of a key observation that eventhough different memory models have different
overall system behaviors, they all share a coherent cache hierarchy. We prove
that this coherent cache hierarchy implements the \emph{store atomicity}
property, which makes it behave like an atomic instantaneous memory with
respect to the rest of the system. We further modularize the proof by defining
what is meant for a processor to be correct independently of the memory system.
As a concrete instance of our framework, we walk through a modular proof of
high-performance multi-processor implementation of Lamport's Sequential
Consistency model. The processors employ \emph{speculative execution} to
generate many concurrent loads to the memory sub-system, while obeying our
correctness criteria. We offer the first formal proof that when an atomic
memory is connected to $n$ speculative processors, the system as a whole is
sequentially consistent.  Additionally, the cache-coherent memory we present in
this paper is the first fully verified system that supports arbitrary cache
hierarchies. Our framework involves describing all our components using labeled
transition systems and we machine-verify all our proofs using the Coq proof
assistant.

### 2014-10-15 Introduction.tex
all the changes

Hardware designers recognize the challenge of implementing such systems, to such an extent that they have
already become some of the most serious real-world adopters of formal methods.
Hardware verification is dominated by model-checking, where abstraction
techniques are used to reduce designs to finite state spaces, which can be
explored exhaustively.  There are limits to the construction of sound
abstractions, so verifications of, e.g., cache-coherence protocols have almost
always treated systems with concrete topologies, involving particular finite
numbers of caches and processors. For instance, explicit-state model checking tools
like Murphi~\cite{murphi} or TLC~\cite{tlc} are only able to handle small and
simplified systems, \eg{} a single-level cache hierarchy with three CPUs, two
addresses, and one bit of data. Symbolic model checking techniques have fared
slightly better; McMillan \etal{} have verified a 2-level MSI protocol based on
the Gigamax distributed multiprocessor using SMV~\cite{gigamax}. Optimizations
on these techniques (\eg{} partial order reduction~\cite{part}, symmetry
reduction~\cite{sym1, sym2}, compositional reasoning~\cite{comp}, \etc{}) scale
the approach further, but they are still unable to handle realistic
hierarchical protocols. We might hope to achieve higher
assurance and understanding of design ideas by verifying \emph{infinite families}
of hardware designs, which resist reduction to finite-state models.  A
potential related benefit is reducing the computational resources required
to formally verify a new design in the family, applying a more general theorem without
much new analysis.

A related benefit of a deductive approach is \emph{modularity}.  It seems to
be rare for hardware practitioners to verify full designs that include
distinct functional components, like processors and memory systems.  One
reason is the algorithmic overhead of exploring such large state spaces.
If we can instead give localized correctness theorems for different
subsystems, then we can compose these theorems in a black-box way to
produce full-system theorems.

The traditional notion of modularity in hardware is closely tied to
\emph{Finite State Machines} (FSMs), more specificially synchronous FSMs. The biggest
problem with using FSM abstraction is that it restricts significantly the kinds of
refinements that can be performed on a module. For example, if we change the
timing of an adder so that it takes 2 clock cycles instead of 1, the whole
system is likely to break. Even though big modules in hardware are often
designed with asynchronous handshaking protocols, current
hardware verification methodologies do not deal with timing refinements.
In this paper, we use the formalism of \emph{labeled transition systems} that
is well-known in process-calculus circles.  It turns out to be exactly the
right way of capturing contracts on interaction between hardware components,
enabling both modular refinement and verification.

%Proofs in this style are quite common in the software world.  The function
%preconditions and postconditions of Hoare logic provide a canonical
%example, and there have been many innovations in modular proof since Hoare's
%original work.  Somehow the hardware verification world has taken less
%advantage of modularity.  In this paper, we propose a new program in
%hardware verification, based on modular proofs using the formalism of
%\emph{labelled transition systems} that is well-known in process-calculus
%circles.  It turns out to be exactly the right way of capturing contracts on
%interaction between hardware components.

Computer architects have intuitions about the contracts of classic
subsystems.  However, these intuitions have not been formalized
in enough detail to enable rigorous proofs.  In carrying out that exercise,
we have also learned some things about architectural design that should
be of interest outside a formal-methods context.

Concretely, we provide a modularization of distributed shared-memory systems
that makes minimal assumptions about networks, provides an abstraction for
memory that allows high-performance implementation of any reasonable memory
model, and allows a full range of realistic performance optimizations in
processors without violating modularity. We want our modular system to permit
\emph{black-box refinement}, that is, nothing beyond the interface contract can
be assumed about other modules.

In each instance we describe a reference model and an implementation model for
a module using labeled transition systems and formally prove that the
implementation is correct (sound) with respect to our reference. For example,
our reference model for memory is monolithic, where each load or store request
generates a (conceptually) instantaneous response, but where requests may be processed
in any order. Such a model is intuitively \emph{store atomic} in the sense that
the effect of a store is visible to all subsequent loads. We then describe a
realistic cache-coherence protocol comprised of a hierarchy of caches, and
formally prove that this system can never produce
a behavior that is not allowed by the reference model. Similarly, we define a
reference model for a processor that executes one instruction at a time and
an implementation model that exploits speculation to increase parallelism,
and we show that the speculative processor implements the reference processor,
in a suitable sense.

In the end, we prove a theorem about an infinite family of hardware systems,
each system having an arbitrary number of speculative processors connected to
a memory with an arbitrary hierarchy of caches.  All our proofs are mechanized
in the Coq proof assistant, and they are modular in the sense of allowing
further optimization to either processor or memory without needing to touch
the proof of the other.  We characterize overall system behavior with a very
simple transition system formalizing \emph{sequential consistency}, and we
conjecture that the methodology generalizes to other popular memory models.

\begin{figure}
\centering
\includegraphics[scale=.43]{zoom}
\caption{Overall system}
\label{zoom}
\end{figure}

Figure \ref{zoom} sketches the sort of hardware system that we have verified.
We show the top-level system design at top right, and we zoom in on boxes standing for the
memory system and a single processor.

Memory is composed of a \emph{hierarchy of caches}, where each cache node (labeled
like ``L1,'' ``L2,'' etc.) communicates only with its neighbors in the graph.
Each processor has a dedicated L1 cache, from which we do our best to satisfy
all memory requests, to avoid the latency of a round-trip with main memory.
With an L1 cache miss, we may still find a hit in a parent cache below the
level of main memory, realizing a smaller but still significant speedup.
We have verified a \emph{directory-based} protocol for coordinating an arbitrary
tree of caches, where each node stores a conservative approximation of its
children's states.

A processor is decomposed into several components.  We have the normal
architectural state, such as values of registers.  Our proofs are generic over
\emph{a family of instruction set architectures}, with parameters for opcode sets and
functions for executing opcodes and decoding them from memory.  Other key
components are a \emph{branch predictor}, which guesses at the control-flow path
that a processor will follow, to facilitate speculation; and a
\emph{reorder buffer (ROB)}, which decides which instructions along that path to
try executing ahead of schedule.  Our proofs apply to an arbitrary branch predictor,
and they work for any reorder buffer satisfying a simple semantic condition.

Figure~\ref{proofs} gives the overall proof structure that we employ to verify
the system of Figure~\ref{zoom}. As we will explain later, $P_\text{so}$
represents the speculative out-of-order processor, $M_c$ the cache memory,
$M_m$ the simple memory, $P_\text{ref}$ a simple decoupled processor, and
SC the simple reference model having multiple processors executing each
instruction atomically thus implementing sequential consistency.  Notation
$A^n$ is for $n$ copies of system $A$ running in parallel.
$\sqsubseteq$ is a refinement operator, capturing a suitable notion of when
one system implements another.  We go into more detail on the symbols from
the figure in the rest of the paper.

Our current framework establishes theorems of the form ``if system $A$ has a run
with some particular observable behavior, then system $B$ also has a run with
the same behavior.''  In this sense, we say that $A$ correctly implements $B$.
Other important properties, such as \emph{deadlock freedom} for $A$ (which
might get stuck without producing any useful behavior), we leave for future
work.

\begin{figure}
\includegraphics[scale=.45]{proofs}
\caption{Overall proof structure}
\label{proofs}
\end{figure}

@@

\begin{itemize}

@@

\item The \emph{first machine-checked proof} of an invalidation-based
  cache-coherence protocol for distributed memories comprised of arbitrary
  cache hierarchies. We believe such a memory is also the correct abstraction for
  implementing relaxed memory models.

\item We use semantics to \emph{elucidate folklore ideas in computer architecture}.
  For instance, we formalize a variety of processor speculation patterns with
  the idea of speculative loads followed by verification loads, and we capture
  a key requirement on memory systems as refinement of a simple transition
  system.

\end{itemize}

As a side effect, we %debunk a common misconception about the semantics of the
have shown a (rather surprising) architectural result about the semantics of
the interface between processor and memory system. One can establish a
sequentially consistent system even if the buffers between the memory system
and the processor do not preserve message order. We prove as much by
establishing simulation of sequential consistency without relying on an
ordering assumption.

\paragraph{Paper Organization:} In Section~\ref{sec:lts} we introduce
our flavor of the labeled transition systems formalism,
including a definition of trace refinement. In
Section~\ref{sec:sc} we give a simple formal model of sequential consistency,
which we decompose further for decoupled shared-memory systems in Section~\ref{sec:ref}.
The following sections refine the two main subcomponents from Figure~\ref{zoom}.
Section~\ref{sec:ooo} discusses definition and verification of a
speculative processor model, and Sections~\ref{sec:cc} and~\ref{sec:ccproof}
respectively define and prove our hierarchical cache-coherence protocol.
Finally, in Section~\ref{sec:finalresult} we show
the whole-system modular proof of our complex system, followed by
discussion of implications in Section~\ref{sec:implications}
and further discussion of related work and other conclusions in
Sections~\ref{relatedWork} and \ref{sec:conclusion}.

The Coq source code of our proofs is included as the non-anonymous
supplement to this paper.

>>>>

The goal of this work is to provide a framework for full verification of such
complex hardware systems. Hardware designers already recognize the challenge of
implementing such systems, to such an extent that they have already become some
of the most serious real-world adopters of formal methods.  Hardware
verification is dominated by model-checking, where abstraction techniques are
used to reduce designs to finite state spaces, which can be explored
exhaustively.  There are limits to the construction of sound abstractions, so
verifications of, e.g., cache-coherence protocols have almost always treated
systems with concrete topologies, involving particular finite numbers of caches
and processors. For instance, explicit-state model checking tools like
Murphi~\cite{murphi} or TLC~\cite{tlc} are only able to handle small and
simplified systems, \eg{} a single-level cache hierarchy with three CPUs, two
addresses, and one bit of data. Symbolic model checking techniques have fared
slightly better; McMillan \etal{} have verified a 2-level MSI protocol based on
the Gigamax distributed multiprocessor using SMV~\cite{gigamax}. Optimizations
on these techniques (\eg{} partial order reduction~\cite{part}, symmetry
reduction~\cite{sym1, sym2}, compositional reasoning~\cite{comp}, \etc{}) scale
the approach further, but they are still unable to handle realistic
hierarchical protocols. We might hope to achieve higher assurance and
understanding of design ideas by verifying \emph{infinite families} of hardware
designs, which resist reduction to finite-state models.  A potential related
benefit is reducing the computational resources required to formally verify a
new design in the family, applying a more general theorem without much new
analysis.

Modularity has long been understood as a key property for effective design and
verification of such complex systems. For the designer, it allows a separation
of concerns increasing robustness by allowing the behavior encapsulated by a
modular boundary to be realized by multiple implementation any of which may be
safely dropped into the system. It also allows a greater measure of
parallelization of design improving development time. Similarly verification is
simplified as modular interface agreements provide a set of natural lemmas to
verify. This leads us towards a decomposition of the whole system verification
task into lemmas of sub-systems, which can be composed in a black-box manner to
produce full-system theorems.

This works introduces an interface enabling true modular implementation and
verification of shared memory systems. We want our modular system to permit
\emph{black-box refinement}, that is, nothing beyond the interface contract can
be assumed about other modules.  That is, we should be able to describe a
modular decomposition where we may replace either the processor nodes, the
memory with locally verified alternatives without changing the correctness of
our whole system.  To this end, we identify a key property called \emph{store
atomicity} shared by the memory component of every memory model. This property
effectively restricts the behavior of the memory sub-system to that of an
\emph{atomic instantaneous memory}. We prove that this property is in fact what
is guaranteed using the cache coherence protocols in a multi-processor system
containing a hierarchy of caches, enabling replacement of the cache hierarchy
with an atomic instantaneous memory. Similarly, on the processor side, we
define what is meant to be a correct single processor. These invariants about
the processor and the memory sub-system remains constant across memory models.
We show a concrete instance of this framework by proving that a highly
speculative multi-processor system connected to a coherent cache hierarchy
implements the sequential consistency memory model.

Our proof methodology involves describing a reference model of a component and
an implementation model for that component using labeled transition systems
(LTS) and formally proving that the implementation is correct (sound) with
respect to our reference. Using LTS for describing hardware has already been
studied \cite{HoeArvind:TRSSynthesis1, Hoe:TCAD}, and there are compilers converting an LTS into efficient
hardware \cite{BSV:LangRef, Bluespec:TFRG}. In fact, a cache coherence protocol implementation in Bluespec was discussed in \cite{DNA:CoherenceImplementation}. In fact, the cache coherence protocol that we
study in this work is obtained from a direct transliteration of Bluespec code
used to implement a multi-processor PowerPC system \cite{Khan:PowerPc}. The
resulting implementation from that Bluespec code is pretty efficient; an 8-core
system can run 55 million instructions overall per second on the BEE FPGA
board, with a 2-level cache hierarchy. All our proofs are mechanized in the Coq
proof assistant, and are included as the non-anonymous supplement to this
paper.

@@

\begin{itemize} 
\item Identification of an interface between the processor and the memory
components of a multi-processor system that enables modular implementation and
verification.

@@

\item The \emph{first machine-checked proof} of an invalidation-based
  cache-coherence protocol for distributed memories comprised of arbitrary
  cache hierarchies. We believe such a memory is also the correct abstraction for
  implementing relaxed memory models.
\end{itemize}

\paragraph{Paper Organization:} In Section~\ref{sec:lts} we introduce our
flavor of the labeled transition systems formalism, including a definition of
trace refinement. In Section~\ref{sec:store-atomicity}, we show a generic
decomposition of any multi-processor system independent of the memory model
that it implements, and discuss the store-atomicity property of the memory
sub-component. In Section~\ref{sec:sc} we
give a simple formal model of sequential consistency.  The
following sections refine the two main subcomponents from Figure~\ref{zoom}.
Section~\ref{sec:ooo} discusses definition and verification of a speculative
processor model, and Sections~\ref{sec:cc} and~\ref{sec:ccproof} respectively
define and prove our hierarchical cache-coherence protocol.  Finally, in
Section~\ref{sec:finalresult} we show the whole-system modular proof of our
complex system and a discussion of related work and other conclusions in
Sections~\ref{relatedWork} and \ref{sec:conclusion}.

%The traditional notion of modularity in hardware is closely tied to
%\emph{Finite State Machines} (FSMs), more specificially synchronous FSMs. The biggest
%problem with using FSM abstraction is that it restricts significantly the kinds of
%refinements that can be performed on a module. For example, if we change the
%timing of an adder so that it takes 2 clock cycles instead of 1, the whole
%system is likely to break. Even though big modules in hardware are often
%designed with asynchronous handshaking protocols, current
%hardware verification methodologies do not deal with timing refinements.
%In this paper, we use the formalism of \emph{labeled transition systems} that
%is well-known in process-calculus circles.  It turns out to be exactly the
%right way of capturing contracts on interaction between hardware components,
%enabling both modular refinement and verification.

%\begin{figure}
%\centering
%\includegraphics[scale=.43]{zoom}
%\caption{Overall system}
%\label{zoom}
%\end{figure}
%
%Figure \ref{zoom} sketches the sort of hardware system that we have verified.
%We show the top-level system design at top right, and we zoom in on boxes standing for the
%memory system and a single processor.
%
%Memory is composed of a \emph{hierarchy of caches}, where each cache node (labeled
%like ``L1,'' ``L2,'' etc.) communicates only with its neighbors in the graph.
%Each processor has a dedicated L1 cache, from which we do our best to satisfy
%all memory requests, to avoid the latency of a round-trip with main memory.
%With an L1 cache miss, we may still find a hit in a parent cache below the
%level of main memory, realizing a smaller but still significant speedup.
%We have verified a \emph{directory-based} protocol for coordinating an arbitrary
%tree of caches, where each node stores a conservative approximation of its
%children's states.
%
%A processor is decomposed into several components.  We have the normal
%architectural state, such as values of registers.  Our proofs are generic over
%\emph{a family of instruction set architectures}, with parameters for opcode sets and
%functions for executing opcodes and decoding them from memory.  Other key
%components are a \emph{branch predictor}, which guesses at the control-flow path
%that a processor will follow, to facilitate speculation; and a
%\emph{reorder buffer (ROB)}, which decides which instructions along that path to
%try executing ahead of schedule.  Our proofs apply to an arbitrary branch predictor,
%and they work for any reorder buffer satisfying a simple semantic condition.
%
%Figure~\ref{proofs} gives the overall proof structure that we employ to verify
%the system of Figure~\ref{zoom}. As we will explain later, $P_\text{so}$
%represents the speculative out-of-order processor, $M_c$ the cache memory,
%$M_m$ the simple memory, $P_\text{ref}$ a simple decoupled processor, and
%SC the simple reference model having multiple processors executing each
%instruction atomically thus implementing sequential consistency.  Notation
%$A^n$ is for $n$ copies of system $A$ running in parallel.
%$\sqsubseteq$ is a refinement operator, capturing a suitable notion of when
%one system implements another.  We go into more detail on the symbols from
%the figure in the rest of the paper.
%
%Our current framework establishes theorems of the form ``if system $A$ has a run
%with some particular observable behavior, then system $B$ also has a run with
%the same behavior.''  In this sense, we say that $A$ correctly implements $B$.
%Other important properties, such as \emph{deadlock freedom} for $A$ (which
%might get stuck without producing any useful behavior), we leave for future
%work.
%
%\begin{figure}
%\includegraphics[scale=.45]{proofs}
%\caption{Overall proof structure}
%\label{proofs}
%\end{figure}

### 2014-10-15 RelatedWork.tex
all the changes

%There has been a lot of work in providing frameworks for axiomatic
%specifications of memory models \cite{alglave2012formal,alglave2014herding}
%and in providing axiomatic memory model specifications for real systems
%through analyis of empirically observed behavior of certain ``litmus tests''
%\cite{mador2012axiomatic,sarkar2012synchronising}. There has also been some
%work in providing operational semantics that describes the observed behavior of
%real hardware \cite{sarkar2011understanding}. Finally, there has been work
%formally relating the axiomatic specification of TSO to operational semantics
%of real x86 and Sparc machines \cite{x86tsocacm10}.
%
%The plethora of work in the area of memory models indeed suggests that this is
%a complex problem, exacerbated by the hardware architect's desire to increase
%hardware performance at the cost of complicating the memory models further. There
%are two aspects where our work can help in tackling this complex problem: (1) we
%are proposing a modular approach for verifying systems in which the components,
%namely the processors and the memory, can be verified against the memory model
%contract independent of each other, and (2) we are also proposing a design
%in which a system obeying sequential consistency is made into a high
%performance machine through the use of speculative loads followed by
%verification. The only downside is waiting for store acknowledgements, but this
%wait does not preclude the machine from performing further computations for
%future instructions without actually committing them. We are hoping to develop
%both these ideas -- of extending our proof approach for more complex memory
%models and designing real hardware which is sequentially consistent but has
%high performance.
%
%
%BELOW IS UNEDITTED FROM OLD PAPER

>>>>

\xxx{TODO: Correct This}

There has been a lot of work in providing frameworks for axiomatic
specifications of memory models \cite{alglave2012formal,alglave2014herding}
and in providing axiomatic memory model specifications for real systems
through analyis of empirically observed behavior of certain ``litmus tests''
\cite{mador2012axiomatic,sarkar2012synchronising}. There has also been some
work in providing operational semantics that describes the observed behavior of
real hardware \cite{sarkar2011understanding}. Finally, there has been work
formally relating the axiomatic specification of TSO to operational semantics
of real x86 and Sparc machines \cite{x86tsocacm10}.

%The plethora of work in the area of memory models indeed suggests that this is
%a complex problem, exacerbated by the hardware architect's desire to increase
%hardware performance at the cost of complicating the memory models further. There
%are two aspects where our work can help in tackling this complex problem: (1) we
%are proposing a modular approach for verifying systems in which the components,
%namely the processors and the memory, can be verified against the memory model
%contract independent of each other, and (2) we are also proposing a design
%in which a system obeying sequential consistency is made into a high
%performance machine through the use of speculative loads followed by
%verification. The only downside is waiting for store acknowledgements, but this
%wait does not preclude the machine from performing further computations for
%future instructions without actually committing them. We are hoping to develop
%both these ideas -- of extending our proof approach for more complex memory
%models and designing real hardware which is sequentially consistent but has
%high performance.
%

%BELOW IS UNEDITTED FROM OLD PAPER
